Toronto's Most Studious Wards:  An Analysis of Annual Toronto Public Library's Circulation and Visits by Wards Prior to and Post-COVID

### 2024-01-25 outputs/paper/paper.qmd
paper upload

---
title: "Toronto's Most Studious Wards:  An Analysis of Annual Toronto Public Library's Circulation and Visits by Wards Prior to and Post-COVID"
subtitle: "Data Visualization in R"
author:
  - Francesca Ye
email: francesca.ye@mail.utoronto.ca
prerequisites: "Datasets taken from opensourcetoronto and can be found here: https://open.toronto.ca/catalogue/"
date: today 
date-format: long
abstract: "Toronto Public Library data sets on branch information, annual circulation, and annual visits were used to examine library usage patterns by ward between 2018 and 2022. Based on the analysis of annual circulation and visits by ward over this five year period, there is evidence that library usage significantly dropped in 2020 when comparing values to 2018 and 2019. Looking beyond the initial outbreak of COVID-19, it also appears that library usage is beginning to recover, although not to pre-pandemic levels, which highlight differences in pre and post-pandemic library usage. It would be beneficial to expand the scope of the analysis to more years before and after 2020 to get a larger sample of library usage patterns in addition to analysis of monthly circulation and visitiation data against a timeline of government mandated COVID-19 measures."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---
```{r}
#| warning: false
#| message: false
#| echo: false

@@

## Introduction
Public libraries have an intrinsic value in communities regardless of if they are used for research, educational, entertainment, or recreational purposes. The availability of public libraries in addition to their material and programming are thought to have an impact on the quality of life in communities. Public libraries can influence a community’s number of jobs, median household income, and level of educational attainment. [@libraryimpact] As a result of the outbreak of COVID-19 which was formally declared a pandemic on March 11, 2020 by the World Health Organization [@who], libraries had their operations affected. Ranging from full closures due to lock downs or social distancing measures which restricted capacity, the landscape upon which public libraries operated rapidly changed. Hamid R. Jamali and Philip Hider’s “Changes in the Usage Pattern of Public Library Collections during the COVID Pandemic” in *Collection Management* looked into the usage patterns of two library networks in Sydney, Australia. They found that physical loans experienced an approximately half million decrease. [@libraryusage]

@@

## Data
Data used in this report were taken from the Open Data Toronto Online Portal [@opendatatoronto]. The specific data sets sourced from Open Data Toronto included *Library Branch General Information* [@librarybranch], *Library Circulation* [@circulationdata], and *Library Visits* [@visitdata]. The data sets were imported into RStudio where the open-source statistical programming language, R [@citeR], was used for cleaning and analysis. The raw data sets were extracted and saved with functionalities of the `readr` [@readr] package. Data cleaning was aided by the use of the `janitor` [@janitor], `tidyverse` [@tidyverse], and `dplyr` [@dplyr] packages. To create tables and graphs for data visualization of the datasets, the `ggplot2` [@ggplot2] and `knitr` [@knitr] packages were used.

### Library Branch General Information
This data set was published by the Toronto Public Library and records the current characteristics of its library branches. The data set is refreshed on an annual basis and includes data regarding branch code, branch name, materiality, address, postal code, website, telephone, square footage, parking spots, workstations, ward number, ward name, present site date, neighbourhood number, neighbourhood name, latitude, longitude, and services offered. [@librarybranch] The data used in this report was captured and accurate as of January 19, 2024.

Basic data cleaning was first applied to the column value names in order to ensure readability. As this is a report focused on investigating annual circulation and visits data by ward and year, it was decided that rows of branches that had a value of `0` in the ‘physical_branch’ which indicated that they did not have a physical location that patrons could enter were removed from the data set. The reason for this removal was to focus on public library branches that were physically accessible to patrons. Additionally, `branch_code`, `ward_name`, and `ward_no` were the only column values that were kept in the cleaned data set. This is because the purpose of this data set was to provide a way for the branch codes present in the *Library Circulation* and *Library Visits* data sets to be assigned their respective ward number for graphical analysis. The final cleaned version of the *Library Branch General Information* the data set has these changes reflected. (@tbl-cleanedbranchdata)

@@

### Library Circulation
This data set was published by the Toronto Public Library and captures the annual circulation of materials at its various branches. The data set is refreshed on an annual basis and includes data regarding the year, branch code, and total circulation. The data used in this report was captured and accurate as of January 19, 2024. According to Open Data Toronto, circulation is counted as any first-time checkouts and renewals of material. This data set also covers physical and electronic materials in its count. Open Data Toronto also notes that circulation totals are a measure of how many of these aforementioned transactions were processed by each respective branch. They also note that online renewals of physical materials are associated with the home branch of the renewing customer. [@circulationdata]

Basic cleaning was first applied to the column value names to ensure readability as a CSV file. Similar to what was done for the cleaning of *Library Branch General Information*, non-physical branches were removed from the data set in addition to the branch code of `VR` which represented electronic circulation totals. This is because the scope of the report focuses only on physical branches that are accessible to patrons. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on a five-year span between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] The aforementioned changes were reflected in the cleaned data set. (@tbl-cleanedcirculationdata)

@@

### Library Visits
This data set was published by the Toronto Public Library and records the annual number of visits at each branch. The data set is refreshed on an annual basis and includes data regarding the year, branch code, and total visits. The data used in this report was captured and accurate as of January 19, 2024. According to Open Data Toronto, visits were measured using sensors at the entrances of library branches and counts were recorded during scheduled operating hours. They note that visit totals were sometimes estimated when sensor issues were occurring. They also state that since visits were based on entries to the branch, the same individuals might be counted multiple times due to re-entrance in one day. [@visitdata]

Basic cleaning was applied to this data set’s column value names to ensure readability. Following what was done on the two previous data sets, non-physical branches were removed from the data set since the scope of the report is focused on physical branches accessible to patrons. Additionally, the `BKONE` branch code was removed since it represented the annual visits of the two bookmobiles. The reason for this removal was due to its lack of fixed physical presence which could be found within one specific ward in Toronto. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on five years between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] These changes to the raw data set are viewable in the cleaned table. (@tbl-cleanedvisitsdata)

@@

### Annual Library Circulation and Visits by Ward
To get data sets that feature the annual circulation and visits total for each ward, the cleaned library branch data set was appended to each respective cleaned data set using the `merge()` function. Data sets were merged by `branch_code` which allowed for the corresponding ward numbers found in the library branch data set to be appended to the rows of annual total data that are sorted by branch code. [@mergedataframes] After the data sets were merged, they were grouped by ward number and then a year to be summarized by total visits or circulation. This process resulted in a data set producing the total annual circulation number for each ward and year between 2018 and 2022 as seen in its table. (@tbl-branchcirculationdata) It also produced a data set producing the total annual visit number for each ward and year between 2018 and 2022 as seen in its resulting table. (@tbl-branchvisitsdata) The reason for creating these data sets was to be able to produce graphs which visualize annual circulation and visit totals by ward instead of branches.

@@

### City-Wide Annual Circulation and Visits
Based on the two previous data sets which aim to display total annual circulation and total by ward, they were also used to create city-wide totals for each year. The data sets were grouped by year and ward totals of circulation and visits were summed. This resulted in one data set which provides the total annual circulation of the Toronto Public Library’s physical and fixed location branches (@tbl-tplannualcirculation) in addition to another data set which provides the total annual visits of the Toronto Public Library’s physical and fixed location branches. (@tbl-tplannualvisits)

@@

## Results
Bar graphs were created with `ggplot2`[@ggplot2] to visualize the cleaned data sets. A bar graph for each year between 2018 and 2022 showing the total circulation by each ward was created. The y-axis scale for all of these graphs started at 0 and ended at 1600000 with increments increasing by 200000 to capture all data values. Additionally, a bar graph for each year between 2018 and 2022 showing the total visits to libraries by each ward was created. The y-axis scale for all of these graphs started at 0 and ended at 2400000 with increments increasing by 200000 to capture all data values.

@@

As seen in a broad view of @fig-circulation2018, @fig-circulation2019, @fig-circulation2020, @fig-circulation2021, and @fig-circulation2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean annual circulation across wards was 910130.2 and 876039.6 respectively. This is comparatively higher than 2020 which has a mean of 	
409095.4. This mean was higher in 2021 at 517534.6 transactions. There was also an increase in 2022 at 620818.8.

@@

Comparatively, Ward 17 in 2018 had a circulation total of 1568603 to Ward 4 in 2020 and 2022’s totals of  811568 and 1190959. This indicates that 2020 did not see many circulation transactions across all wards by looking at the highest values. By looking at the highest values of these years, there is evidence the annual circulation total by ward hit its trough in 2020 after a year-over-year decline starting in 2018 and began to recover from then onward.

@@

Ward 11 had visit totals of 2349923 in 2018, 435484 in 2021, and 1163179 in 2022. By viewing the highest ward total for each year, it can be concluded that visitation decreased at a peak in 2021 across all wards. By looking at the highest values of 2018, 2020, and 2020, there is evidence the annual visit total by ward hit its trough in 2021.

### City-Wide Annual Circulation and Visits
Tables that provide a summary of the city-wide total visits (@tbl-tplannualvisits) and total circulation (@tbl-tplannualcirculation) for each year. Annual circulation and visitation metrics have high standard deviations of 5502796 and 6349144, respectively. The high standard deviations imply that the overall annual variance of these metrics is high, therefore indicating values are spread out from the mean. This is seen in the graphs for both usage metrics as 2020 and 2021 reveal significantly smaller values compared to 2018 and 2019. By viewing the annual data, it is clear that the graphs represent the annual data after comparing the table to the aforementioned sections.

@@

Limitations also exist in the analysis of the data sets in this report. While this report does analyze and graph the annual ward totals of library circulation and visits, expanding the chronological scope of the report could provide greater insight into whether or not post-2020 data was directly affected by the pandemic and not other factors. The report organizes data by ward and therefore it might have been beneficial to compare total circulation and visits to demographic ward data to rule out other factors aside from the pandemic as causes of the severe decrease seen in 2020. Annual data is also used for the report and it is worth investigating library usage metrics against periods of lockdown or mandated social distancing on a daily or monthly basis across the same five-year period. However, monthly and daily data is not available on Open Data Toronto [@opendatatoronto] This could further support the assertion that there were changes in total circulation and visits that can be categorized as pre and post-pandemic.

>>>>

---
title: "Toronto's Most Studious Wards:  An Analysis of Annual Toronto Public Library's Circulation and Visits by Wards Prior to and Post-COVID"
subtitle: "Data Visualization in R"
author:
  - Francesca Ye
email: francesca.ye@mail.utoronto.ca
prerequisites: "Datasets taken from opensourcetoronto and can be found here: https://open.toronto.ca/catalogue/"
date: today 
date-format: long
abstract: "Toronto Public Library data sets on branch information, annual circulation, and annual visits were used to examine library usage patterns by ward between 2018 and 2022. Based on the analysis of annual circulation and visits by ward over this five year period, there is evidence that library usage significantly dropped in 2020 when comparing values to 2018 and 2019. Looking beyond the initial outbreak of COVID-19, it also appears that library usage is beginning to recover, although not to pre-pandemic levels, which highlight differences in pre and post-pandemic library usage. It would be beneficial to expand the scope of the analysis to more years before and after 2020 to get a larger sample of library usage patterns in addition to analysis of monthly circulation and visitiation data against a timeline of government mandated COVID-19 measures."
thanks: "The full code and data used in this paper is available at: https://github.com/Francesca-Ye/tpl_activity_analysis"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---
```{r}
#| warning: false
#| message: false
#| echo: false

@@

## Introduction
Public libraries have an intrinsic value in communities regardless of if they are used for research, educational, entertainment, or recreational purposes. The availability of public libraries in addition to their material and programming are thought to have an impact on the quality of life in communities. Public libraries can influence a community’s number of jobs, median household income, and level of educational attainment [@libraryimpact]. As a result of the outbreak of COVID-19 which was formally declared a pandemic on March 11, 2020 by the World Health Organization [@who], libraries had their operations affected. Ranging from full closures due to lock downs or social distancing measures which restricted capacity, the landscape upon which public libraries operated rapidly changed. Hamid R. Jamali and Philip Hider’s “Changes in the Usage Pattern of Public Library Collections during the COVID Pandemic” in *Collection Management* looked into the usage patterns of two library networks in Sydney, Australia. They found that physical loans experienced an approximately half million decrease [@libraryusage].

@@

## Data
Data used in this report were taken from the Open Data Toronto Online Portal [@opendatatoronto]. The specific data sets sourced from Open Data Toronto included *Library Branch General Information* [@librarybranch], *Library Circulation* [@circulationdata], and *Library Visits* [@visitdata]. The data sets were imported into RStudio where the open-source statistical programming language, R [@citeR], was used for cleaning and analysis. The raw data sets were extracted and saved with functionalities of the `readr` [@readr] package. Data cleaning was aided by the use of the `janitor` [@janitor], `tidyverse` [@tidyverse], and `dplyr` [@dplyr] packages. To create tables and graphs for data visualization of the datasets, the `ggplot2` [@tidyverse] and `knitr` [@knitr] packages were used.

### Library Branch General Information
This data set was published by the Toronto Public Library and records the current characteristics of its library branches. The data set is refreshed on an annual basis and includes data regarding branch code, branch name, materiality, address, postal code, website, telephone, square footage, parking spots, workstations, ward number, ward name, present site date, neighbourhood number, neighbourhood name, latitude, longitude, and services offered [@librarybranch]. The data used in this report was captured and accurate as of January 19, 2024.

Basic data cleaning was first applied to the column value names in order to ensure readability. As this is a report focused on investigating annual circulation and visits data by ward and year, it was decided that rows of branches that had a value of `0` in the ‘physical_branch’ which indicated that they did not have a physical location that patrons could enter were removed from the data set. The reason for this removal was to focus on public library branches that were physically accessible to patrons. Additionally, `branch_code`, `ward_name`, and `ward_no` were the only column values that were kept in the cleaned data set. This is because the purpose of this data set was to provide a way for the branch codes present in the *Library Circulation* and *Library Visits* data sets to be assigned their respective ward number for graphical analysis. The final cleaned version of the *Library Branch General Information* the data set has these changes reflected (@tbl-cleanedbranchdata).

@@

### Library Circulation
This data set was published by the Toronto Public Library and captures the annual circulation of materials at its various branches. The data set is refreshed on an annual basis and includes data regarding the year, branch code, and total circulation. The data used in this report was captured and accurate as of January 19, 2024. According to Open Data Toronto, circulation is counted as any first-time checkouts and renewals of material. This data set also covers physical and electronic materials in its count. Open Data Toronto also notes that circulation totals are a measure of how many of these aforementioned transactions were processed by each respective branch. They also note that online renewals of physical materials are associated with the home branch of the renewing customer [@circulationdata].

Basic cleaning was first applied to the column value names to ensure readability as a CSV file. Similar to what was done for the cleaning of *Library Branch General Information*, non-physical branches were removed from the data set in addition to the branch code of `VR` which represented electronic circulation totals. This is because the scope of the report focuses only on physical branches that are accessible to patrons. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on a five-year span between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] The aforementioned changes were reflected in the cleaned data set (@tbl-cleanedcirculationdata).

@@

### Library Visits
This data set was published by the Toronto Public Library and records the annual number of visits at each branch. The data set is refreshed on an annual basis and includes data regarding the year, branch code, and total visits. The data used in this report was captured and accurate as of January 19, 2024. According to Open Data Toronto, visits were measured using sensors at the entrances of library branches and counts were recorded during scheduled operating hours. They note that visit totals were sometimes estimated when sensor issues were occurring. They also state that since visits were based on entries to the branch, the same individuals might be counted multiple times due to re-entrance in one day [@visitdata].

Basic cleaning was applied to this data set’s column value names to ensure readability. Following what was done on the two previous data sets, non-physical branches were removed from the data set since the scope of the report is focused on physical branches accessible to patrons. Additionally, the `BKONE` branch code was removed since it represented the annual visits of the two bookmobiles. The reason for this removal was due to its lack of fixed physical presence which could be found within one specific ward in Toronto. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on five years between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values [@subsetfunction]. These changes to the raw data set are viewable in the cleaned table (@tbl-cleanedvisitsdata).

@@

### Annual Library Circulation and Visits by Ward
To get data sets that feature the annual circulation and visits total for each ward, the cleaned library branch data set was appended to each respective cleaned data set using the `merge()` function. Data sets were merged by `branch_code` which allowed for the corresponding ward numbers found in the library branch data set to be appended to the rows of annual total data that are sorted by branch code. [@mergedataframes] After the data sets were merged, they were grouped by ward number and then a year to be summarized by total visits or circulation. This process resulted in a data set producing the total annual circulation number for each ward and year between 2018 and 2022 as seen in its table (@tbl-branchcirculationdata). It also produced a data set producing the total annual visit number for each ward and year between 2018 and 2022 as seen in its resulting table (@tbl-branchvisitsdata). The reason for creating these data sets was to be able to produce graphs which visualize annual circulation and visit totals by ward instead of branches.

@@

### City-Wide Annual Circulation and Visits
Based on the two previous data sets which aim to display total annual circulation and total by ward, they were also used to create city-wide totals for each year. The data sets were grouped by year and ward totals of circulation and visits were summed. This resulted in one data set which provides the total annual circulation of the Toronto Public Library’s physical and fixed location branches (@tbl-tplannualcirculation) in addition to another data set which provides the total annual visits of the Toronto Public Library’s physical and fixed location branches (@tbl-tplannualvisits).

@@

## Results
Bar graphs were created with `ggplot2`[@tidyverse] to visualize the cleaned data sets. A bar graph for each year between 2018 and 2022 showing the total circulation by each ward was created. The y-axis scale for all of these graphs started at 0 and ended at 1,600,000 with increments increasing by 200,000 to capture all data values. Additionally, a bar graph for each year between 2018 and 2022 showing the total visits to libraries by each ward was created. The y-axis scale for all of these graphs started at 0 and ended at 2,400,000 with increments increasing by 200,000 to capture all data values.

@@

As seen when broadly viewing @fig-circulation2018, @fig-circulation2019, @fig-circulation2020, @fig-circulation2021, and @fig-circulation2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean annual circulation across wards was 910,130.2 and 876,039.6 respectively. This is comparatively higher than 2020 which has a mean of 409,095.4. This mean was higher in 2021 at 517,534.6 transactions. There was also an increase in 2022 at 620,818.8.

@@

Comparatively, Ward 17 in 2018 had a circulation total of 1,568,603 to Ward 4 in 2020 and 2022’s totals of  811,568 and 1,190,959. This indicates that 2020 did not see many circulation transactions across all wards by looking at the highest values. By looking at the highest values of these years, there is evidence the annual circulation total by ward hit its trough in 2020 after a year-over-year decline starting in 2018 and began to recover from then onward.

@@

Ward 11 had visit totals of 2,349,923 in 2018, 435,484 in 2021, and 1,163,179 in 2022. By viewing the highest ward total for each year, it can be concluded that visitation decreased at a peak in 2021 across all wards. By looking at the highest values of 2018, 2020, and 2020, there is evidence the annual visit total by ward hit its trough in 2021.

### City-Wide Annual Circulation and Visits
Tables that provide a summary of the city-wide total visits (@tbl-tplannualvisits) and total circulation (@tbl-tplannualcirculation) for each year. Annual circulation and visitation metrics have high standard deviations of 5,502,796 and 6,349,144, respectively. The high standard deviations imply that the overall annual variance of these metrics is high, therefore indicating values are spread out from the mean. This is seen in the graphs for both usage metrics as 2020 and 2021 reveal significantly smaller values compared to 2018 and 2019. By viewing the annual data, it is clear that the graphs represent the annual data after comparing the table to the aforementioned sections.

@@

Limitations also exist in the analysis of the data sets in this report. While this report does analyze and graph the annual ward totals of library circulation and visits, expanding the chronological scope of the report could provide greater insight into whether or not post-2020 data was directly affected by the pandemic and not other factors. The report organizes data by ward and therefore it might have been beneficial to compare total circulation and visits to demographic ward data to rule out other factors aside from the pandemic as causes of the severe decrease seen in 2020. Annual data is also used for the report and it is worth investigating library usage metrics against periods of lockdown or mandated social distancing on a daily or monthly basis across the same five-year period. However, monthly and daily data is not available on Open Data Toronto [@opendatatoronto]. This could further support the assertion that there were changes in total circulation and visits that can be categorized as pre and post-pandemic.
